---
title: "From stable demographic profiles to biomass reconstruction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From stable demographic profiles to biomass reconstruction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`StableBiomass` extends age-structured stable demographic profiles into
biomass-based quantities. It is designed to work naturally with outputs
derived from `StablePopulation`.

The basic workflow is:

```text
fertility rates + Weibull beta + body mass by age
  -> stable survivorship and demographic profile
  -> live biomass by age
  -> mortality biomass by age
  -> birth biomass by age
  -> summaries by species, age class, and body-mass class
```

# Input data

We start with an age-specific fertility schedule and a body-mass table.

```{r}
library(StableBiomass)

fertility_rates <- c(0, 0, 0.30, 0.75, 0.60, 0.20)

body_mass <- data.frame(
  age = 0:5,
  body_mass = c(5, 20, 45, 70, 80, 82)
)

body_mass
```

The fertility vector represents age-specific fertility values, `m_x`.
The body-mass table gives the expected body mass for each age class.

# From fertility to biomass

The function `derive_biomass_from_fertility()` connects the demographic
reconstruction step with the biomass step.

```{r}
result <- derive_biomass_from_fertility(
  fertility_rates = fertility_rates,
  beta = 1.2,
  body_mass = body_mass
)

names(result)
```

The returned object contains three components:

- `reconstruction`: Weibull parameters and reconstructed survivorship.
- `demographic_profile`: the demographic table used for biomass calculations.
- `biomass_profile`: the biomass profile by age class.

```{r}
result$demographic_profile$table
```

# Biomass profile by age

The biomass profile combines the stable demographic quantities with
age-specific body mass.

```{r}
biomass_profile <- result$biomass_profile
biomass_profile
```

The main biomass quantities are:

- `live_biomass = R * body_mass`
- `mortality_biomass = D * body_mass`
- `birth_biomass = B * body_mass`

# Species-level summary

`summarise_biomass_profile()` converts the age-structured table into
species-level quantities.

```{r}
biomass_summary <- summarise_biomass_profile(biomass_profile)
biomass_summary
```

This summary is useful for comparing species or for reporting the main
biomass outputs in a compact table.

# Age-class summaries

Age classes can be defined by the user. For example, the following table
groups ages into juvenile, subadult, and adult classes.

```{r}
age_classes <- data.frame(
  age_class = c("juvenile", "subadult", "adult"),
  min_age = c(0, 2, 4),
  max_age = c(1, 3, 5)
)

age_classes
```

The function `summarise_age_classes()` aggregates demographic and biomass
quantities by these age classes.

```{r}
age_class_summary <- summarise_age_classes(
  biomass_profile = biomass_profile,
  age_classes = age_classes
)

age_class_summary
```

# Body-mass classes

For community and trophic applications, body-mass classes provide a bridge
between age-structured profiles and prey-size categories.

```{r}
mass_classes <- data.frame(
  mass_class = c("small", "medium", "large"),
  min_mass = c(0, 25, 75),
  max_mass = c(25, 75, Inf)
)

classified_profile <- classify_body_mass(
  data = biomass_profile,
  mass_classes = mass_classes
)

classified_profile
```

This classified profile prepares the data for later community-level
functions, such as total biomass output and available biomass by prey-size
class.

# Excel workflow

The same species-level workflow can be run from an Excel file using
`run_species_biomass_excel()`.

The input file must contain these columns:

```text
age | mx | body_mass | beta
```

The `beta` column can contain one value in the first row and empty cells
below it.

Example input:

```text
age | mx   | body_mass | beta
0   | 0.00 | 5         | 1.2
1   | 0.00 | 20        |
2   | 0.30 | 45        |
3   | 0.75 | 70        |
4   | 0.60 | 80        |
5   | 0.20 | 82        |
```

The Excel workflow is called as follows:

```{r, eval = FALSE}
run_species_biomass_excel(
  input_file = "species_input.xlsx",
  output_file = "species_biomass_output.xlsx"
)
```

The output workbook contains demographic profiles, biomass profiles,
species-level summaries, metadata, and optionally age-class summaries.

# Interpretation

`StableBiomass` translates stable demographic structure into biomass-based
quantities. The central interpretation is:

```text
R_x * body_mass_x -> live biomass by age
D_x * body_mass_x -> mortality biomass by age
B_x * body_mass_x -> birth biomass by age
```

This provides the foundation for later community-level analyses, where
mortality biomass and prey-size classes can be used to estimate biomass
availability across species, sites, and ecological scenarios.
